Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-09 paper/paper.qmd
update bibliography and start move toward complete draft

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

There are many aspects of trustworthy data science, including education, culture, and workflow to name a few. One aspect of special importance is the tools that we use to do data science: the computers, programming languages, and environments. The wide release of user-friendly Large Language Models (LLMs), particularly OpenAI's ChatGPT, is a rare instance of a new and powerful tool being made available to add to a data scientist's toolbelt.

Like all new tools, however, LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into mainstream conversation, quickly being put to use in both industry and academia. By now, many people, especially educators and students, have some experience with LLMs, in both personal and professional contexts.

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@Winkler_2018]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Okaibedi_2023], the development of critical thinking skills, and what effective learning looks like [@BaidooAnu_2023].

@@

At the high school level @Lazar_2023 conducted a informal survey of secondary school teachers and students on their opinions of ChatGPT, and found that while LLMs could: help creativity, provide academic support when teachers were unavailable, and model certain types of writing well; teachers were also aware of LLMs potential to limit students' learning in certain ways through over-reliance. Beyond academic integrity concerns, teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward and usage of AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% 'strongly agreed' that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. In particular, respondents found that using it to write whole papers as inappropriate, while using it for basic tasks like general assistance, writing feedback and basic data visualization was perceived more appropriately.

@@

# Writing
writing_perceptions_long <-
  all_data |>
  select(
    writing_is_easy_for_me,
    i_like_to_write,
    i_believe_it_is_important_to_be_a_good_writer,
    when_i_edit_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_writing,
    i_am_confident_in_my_overall_writing_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "writing_is_easy_for_me" ~ "\"Writing is easy for me.\"",
      statements == "i_like_to_write" ~ "\"I like to write.\"",
      statements == "i_believe_it_is_important_to_be_a_good_writer" ~ "\"I believe it is important\nto be a good writer.\"",
      statements == "when_i_edit_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I edit it is easy for\nme to catch my mistakes.\"",
      statements == "i_feel_confident_sharing_my_writing" ~ "\"I feel confident\nsharing my writing.\"",
      statements == "i_am_confident_in_my_overall_writing_ability" ~ "\"I am confident in my\noverall writing ability.\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different from me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

@@

# Coding
coding_perceptions_long <-
  all_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder,
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      statements == "i_like_to_code" ~ "\"I like to code\"",
      statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my\nmistakes\"",
      statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident\n sharing my code\"",
      statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different from me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

@@

Those README statements were gathered and parsed using OpenAI's ChatGPT 4o model (as at 26 July). The following prompt was used: 'The following statement is about to what extent LLMs were used by a student. Please characterize it as one of: "None", "Minimal", "Somewhat", "Extensive", "Unsure". Respond with only one of those options.' All classifications were then manually checked by hand for reasonableness.

@@

The third, and final, component is the mark, in percentages, on the final paper (@fig-mark). The overall mean was 78% and standard deviation was 17 percentage points. However there was considerable differences by the extent of LLM usage. For extensive use, the mean was 82% and the standard deviation was 15 percentage points. For somewhat usage, the mean was 77% and the standard deviation was 16 percentage points. For minimal usage, the mean was 74% and the standard deviation was 23 percentage points. Finally for no usage, the mean was 75% and the standard deviation was 19 percentage points.

```{r fig-mark}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Mark on final paper, in percentages, and LLM usage in final paper

# all_data |>
#   group_by(llm_usage) |>
#   summarize(mean = mean(mark) |> round(2),
#             sd = sd(mark) |> round(2))
#
# all_data |>
#   summarize(mean = mean(mark) |> round(2),
#             sd = sd(mark) |> round(2))

all_data |>
  ggplot(mapping = aes(x = mark, color = )) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  facet_wrap(vars(llm_usage),
             ncol = 2,
             labeller = labeller(
               llm_usage = c(
                 "Extensive" = "Extensive LLM usage",
                 "Minimal" = "Minimal LLM usage",
                 "None" = "No LLM usage",
                 "Somewhat" = "Some LLM usage"
               )
             )) +
  theme_minimal()
```

@@

The goal of our modelling strategy is to better understand how respondents result on their final paper is associated with their self-reported LLM usage. The result on the final paper is a proportion and some respondents got full marks and so we use zero-one-inflated beta regression. Here we briefly describe the model that we use, which follows @kurztotherescue. Diagnostics are included in [Appendix -@sec-model-details].

Define $y_i$ as the percentage received on the final paper. Then $\beta_1$ is the characterization of self-reported LLM usage and $\beta_2$ is self-reported GPA.

\begin{align}
y_i &\sim \mbox{Beta}(\mu_i, \phi) \\
logit(\mu_i) &= \beta_0 + \beta_1 \times \mbox{LLM usage}_i + \beta_2 \times \mbox{GPA}_i\\
\beta_0, \beta_1, \beta_2  &\sim \mbox{Normal}(0, 1) \\
\phi &\sim \mbox{Gamma}(4, 0.1)
\end{align}

We estimate the model in `R` [@citeR] using `brms` [@brms].

The expected relationship between LLM usage and final mark is unclear. It may be that stronger students used LLMs in a more sophisticated way or that they did not need to use them. However, the relationship between GPA and final mark is expected to be positive.

@@

```{r fig-gpamodeldiagnostics}
#| fig-height: 7
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Model including self-reported GPA diagnostics

>>>>

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

There are many aspects of trustworthy data science, including education, culture, and workflow to name a few. One aspect of special importance is the tools that we use to do data science: the computers, programming languages, and environments among other factors. The wide release of user-friendly Large Language Models (LLMs), particularly OpenAI's ChatGPT, is the rare instance of a new tool being added to the data science toolkit.

Like all new tools, however, LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. LLMs have quickly been put to use in both industry and academia. By now, many people, especially educators and students, have some experience with LLMs, in both personal and professional contexts.

In the context of teaching statistics and data science, LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots to act as personalized tutors for students [@Winkler_2018]. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity [@Eke2023], the development of critical thinking skills, and what effective learning looks like [@BaidooAnu_2023].

@@

At the high school level @Lazar_2023 conducted an informal survey of secondary school teachers and students on their opinions of ChatGPT, and found that while LLMs could: help creativity, provide academic support when teachers were unavailable, and model certain types of writing well; teachers were also aware of LLMs potential to limit students' learning in certain ways through over-reliance. Beyond academic integrity concerns, teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward and usage of AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% "strongly agreed" that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. In particular, respondents found that using it to write whole papers as inappropriate, while using it for basic tasks like general assistance, writing feedback and basic data visualization was perceived more appropriately.

@@

# Writing
writing_perceptions_long <-
  all_data |>
  select(
    writing_is_easy_for_me,
    i_like_to_write,
    i_believe_it_is_important_to_be_a_good_writer,
    when_i_edit_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_writing,
    i_am_confident_in_my_overall_writing_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "writing_is_easy_for_me" ~ "\"Writing is easy for me.\"",
      statements == "i_like_to_write" ~ "\"I like to write.\"",
      statements == "i_believe_it_is_important_to_be_a_good_writer" ~ "\"I believe it is important\nto be a good writer.\"",
      statements == "when_i_edit_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I edit it is easy for\nme to catch my mistakes.\"",
      statements == "i_feel_confident_sharing_my_writing" ~ "\"I feel confident\nsharing my writing.\"",
      statements == "i_am_confident_in_my_overall_writing_ability" ~ "\"I am confident in my\noverall writing ability.\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different\nfrom me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

@@

# Coding
coding_perceptions_long <-
  all_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder,
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  ) |>
  pivot_longer(cols = everything(),
               names_to = "statements",
               values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      statements == "i_like_to_code" ~ "\"I like to code\"",
      statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my\nmistakes\"",
      statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident\n sharing my code\"",
      statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different\nfrom me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

@@

Those README statements were gathered and parsed using OpenAI's ChatGPT 4o model (as at 26 July). The following prompt was used:

> The following statement is about to what extent LLMs were used by a student. Please characterize it as one of: "None", "Minimal", "Somewhat", "Extensive", "Unsure". Respond with only one of those options.

All classifications were then manually checked by hand for reasonableness.

@@

The third, and final, component is the mark, in percentages, on the final paper (@fig-mark-1). The overall mean was 78% and standard deviation was 17 percentage points. However there was considerable differences by the extent of LLM usage (@tbl-mark; @fig-mark-2).

```{r fig-mark}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of marks on final paper (in percentages)
#| fig-subcap: ["Overall", "By self-reported LLM usage"]

all_data |>
  select(mark) |>
  ggplot(aes(x = mark)) +
  geom_histogram(binwidth = 0.05) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  theme_minimal()

all_data |>
  ggplot(mapping = aes(x = mark)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  facet_wrap(vars(llm_usage),
             ncol = 2,
             labeller = labeller(
               llm_usage = c(
                 "Extensive" = "Extensive LLM usage",
                 "Minimal" = "Minimal LLM usage",
                 "None" = "No LLM usage",
                 "Somewhat" = "Some LLM usage"
               )
             )) +
  theme_minimal()
```

```{r tbl-mark}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Self-reported LLM usage in final paper and final paper mark

all_data |>
  group_by(llm_usage) |>
  mutate(llm_usage = factor(llm_usage, levels = c(
    "Extensive", "Somewhat", "Minimal", "None"
  ))) |>
  summarize(mean = mean(mark) |> round(2),
            sd = sd(mark) |> round(2)) |>
  kable(col.names = c("Self-reported LLM usage", "Mean", "Std dev"))
```

@@

The goal of our modelling strategy is to better understand how a respondents' result on their final paper associates with their self-reported LLM usage.  Students received their result on the final paper as a proportion (a 0-100% mark), so we use zero-one-inflated beta regression to best fit the distribution in this interval. We use this type of regression given the ability of the beta distribution to effectively describe continuous and skewed or heteroskedastic datasets, which we can see in the asymmetrical histogram of the distribution of marks.

Transforming the regression with a logit function as the link function ensures that the predicted values stay between zero and one, respecting the boundaries of the data. Some students got full marks for the paper, so we use the zero-one-inflated beta model.

Here we briefly describe the model that we use, which follows @kurztotherescue. Model diagnostics are included in [Appendix -@sec-model-details].

Define $y_i$ as the percentage received on the final paper. Then $\beta_1$ is the characterization of self-reported LLM usage and $\beta_2$ is self-reported GPA.
\begin{align}
y_i &\sim \mbox{Beta}(\mu_i, \phi) \\
logit(\mu_i) &= \beta_0 + \beta_1 \times \mbox{LLM usage}_i + \beta_2 \times \mbox{GPA}_i\\
\beta_0, \beta_1, \beta_2  &\sim \mbox{Normal}(0, 1) \\
\phi &\sim \mbox{Gamma}(4, 0.1)
\end{align}

We estimate the model in `R` [@citeR] as well, using `brms` [@brms].

The expected relationship between LLM usage and final mark is unclear. It could be that stronger students used LLMs in a more sophisticated way or that they did not need to use them. However, the relationship between GPA and final mark is expected to be positive.

@@

```{r fig-gpamodeldiagnostics}
#| fig-height: 6
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Model including self-reported GPA diagnostics
